---
title: "Curating data for 'high-confidence' heteroplasmies"
author: "Arslan Zaidi"
date: "11/13/2019"
output: pdf_document
---

# Introduction

Here, we curated our dataset (list of heteroplasmies and list of individuals and families to be analyzed) based on a number of criteria as listed in Fig. S3. Briefly, we wanted to: 

-  keep individuals with both tissues (blood+buccal) sequenced
-  keep families with at least one mother and two children sequenced
-  keep only one replicate for samples that were sequenced more than once
-  remove heteroplasmies that we flagged as being 'suspicious'

# Methodological details of filtering strategy

Load necessary libraries.

```{r echo=T,warning=FALSE,results=FALSE}

library(data.table)
library(dplyr)
library(ggplot2)
library(here)

```

Load 'fam file' - file containing information on each sample (which family they belong to, their age, which run they were sequenced in etc.), load heteroplasmy table, and merge these two tables together.

```{r}

#load fam file
fam<-fread(here("Data_files/famfile_09272018.txt"),header=T,sep="\t")

#remove rows corresponding to runs 13, 13n, 29, and 29n
#these are replicate runs which are now merged into a single record
fam<-fam[-which(fam$run%in%c("TR13n","TR29n","TR13","TR29")),]

#load heteroplasmy table
hets<-fread(here("Data_files/Allruns_2_33_08222018.header.trim.rd.hets"),header=T,sep="\t")

#starting number of heteroplasmies: 1473

#merge hets with fam file
hets2<-merge(hets,fam,by="fqid",all.x=T)

```

Some rows/heteroplasmies (N = 9) were not merged with fam file - fam file does not have information on these samples. Figure out why we don't have info on these and remove them.

```{r}

hets2[which(is.na(hets2$FID)),'fqid']

```

These samples were sequenced for the Rebolledo et al. 2014 [@Rebolledo 2014] but were removed for this project becasue they come from families from which a second child was not sequenced. Remove heteroplasmies, if any, from these samples.

```{r}

hets2<-hets2[which(is.na(hets2$FID)=="FALSE"),]

#remaining hets: 1473

```

Check if all families are represented in the heteroplasmy file - i.e. does every family have at least one heteroplasmy?

```{r}

unique(fam$FID)[-which(unique(fam$FID)%in%unique(hets2$FID))]

#F326 does not have detectable heteroplasmies

```

Keep one of any samples that were sequenced more than once.

```{r}

#load list of duplicates sequenced before run 33 (TR33)
dups<-fread(here("Data_files/duplicates_removed_08092018.txt"),header=F)

#load duplicates of samples sequenced in run 33 (TR33)
dups.33<-fread(here("Data_files/TR33_duplicates.txt"))

#remove duplicated sequences from heteroplasmy file
hets3<-hets2[-which(hets2$fqid%in%dups$V1),]

#881 heteroplasmies left after removing duplicates

#remove older samples sequenced in TR33
hets3<-hets3[-which(hets3$fqid%in%dups.33$fqid_old),]

#797 heteroplasmies remaining

#remove duplicated samples from fam file as well 
fam2<-fam[-which(fam$fqid%in%dups$V1|fam$fqid%in%dups.33$fqid_old),]

```

After these filters, check how many tissues sequenced per person. This is mostly to make sure that after these filters, there aren't any samples for which only have one tissue sequenced. Remove such samples from further analysis.

```{r}

dfam.2t<-fam2%>%
  group_by(individual_id)%>%
  summarize(ntissue=length(unique(tissue_id)))

# > dfam.2t[which(dfam.2t$ntissue<2),]
# individual_id ntissue
# 70         F171g1       1
# 161        F239g1       1
# 185        F254g1       1
# 188    F254g1m1c2       1
# 202      F261m1c1       1
# 291      F308m1c2       1
# 329      F330m1c1       1
# 332      F332m1c1       1

#remove these individuals from heteroplasmy file
hets4<-hets3[-which(hets3$individual_id%in%c("F171g1",
                                            "F239g1",
                                            "F254g1",
                                            "F254g1m1c2",
                                            "F261m1c1",
                                            "F308m1c2",
                                            "F330m1c2",
                                            "F332m1c1")),]

#789 heteroplasmies remaining

#remove from fam file as well
fam3<-fam2[-which(fam2$individual_id%in%c("F171g1",
                                          "F239g1",
                                          "F254g1",
                                          "F254g1m1c2",
                                          "F261m1c1",
                                          "F308m1c2",
                                          "F330m1c2",
                                          "F332m1c1")),]


```


Now check to make sure each mother has at least two children. Keep only families where we have at least one mother and two children.

```{r}

#how many mothers have 2 children
dfam.2c<-fam3%>%
  group_by(mother_id,fam_str)%>%
  summarize(nchildren=length(unique(individual_id)))


#dfam.2c[which(dfam.2c$nchildren<2),]
#mothers with less than two children
# mother_id    fam_str nchildren
# 1     F105g1    0-1-1-2         1
# 2     F113g1    0-0-1-2         1
# 3     F116g1    0-1-1-2         1
# 4     F140g1    0-1-1-2         1
# 5     F171g1    0-1-1-2         1
# 6    F172tg1    1-1-1-3         1
# 7     F239g1 0-1-1-3-1t         1
# 8    F249tg1    1-2-1-2         1
# 9     F254g1    0-1-1-2         1
# 10  F254g1m1    0-1-1-2         1
# 11    F261m1    0-0-1-2         1
# 12    F289g1    0-1-1-2         1
# 13    F291g1    0-1-1-3         1
# 14    F297g1    0-1-1-2         1
# 15    F330m1    0-0-1-2         1
# 16    F332m1    0-0-1-2         1

#remove these families
fam4<-fam3[-which(fam3$FID%in%c("F254","F261","F330","F332")),]
hets5<-hets4[-which(hets4$FID%in%c("F254","F261","F330","F332"))]

#775 heteroplasmies remaining

```

Each family has a 'family structure' associated with it, which details which of the pedigree structures listed in Fig. S1 the family belongs to. This information is helpful in downstream analyses because of the highly nested organization of samples. If the removal of a sample or an individual from a family changes the structure of a family, edit this in the fam file. 

```{r}

#change fam structure and mother category for these families

fam4$fam_cat[which(fam4$FID=="F171")]<-"m1c2"
fam4$fam_str[which(fam4$FID=="F171")]<-"0-0-1-2"

fam4$fam_cat[which(fam4$FID=="F239")]<-"m1c3"
fam4$fam_str[which(fam4$FID=="F239")]<-"0-0-1-3-1t"

fam4$fam_cat[which(fam4$FID=="F308")]<-"m1c2"
fam4$fam_str[which(fam4$FID=="F239")]<-"0-0-1-2"

#F308 initially had three children. Since we removed the 2nd child, we have to relabel the 3rd child as c2 under "levels" column
fam4$level[which(fam4$individual_id=="F308m1c3")]<-"c2"

#write the 'cleaned' fam file to file
fwrite(fam4,here("Data_files/famfile_cleared_conservative_09272018.txt"),sep="\t",col.names=T,row.names=F,quote=F)

```

Clean up the count file accordingly. The count file contains the allele frequency information of each identified heteroplasmy from all family members (even those who may not be carrying the heteroplasmy). Remove the heteroplasmies/samples we removed earlier from the count file. 

```{r}

#### load count file for all family frequencies ####
counts<-fread("https://www.dropbox.com/s/gyk54cgek4kooo4/Allruns_08222018.trim.rd.counts?dl=1",header=F,sep="\t")

colnames(counts)<-c('fqid','ref','position','A','C','G','T','a','c','g','t','cvrg','nalleles','major','minor','maf')

#merge with fam file
counts2<-merge(counts,fam4,by="fqid",all.x=T)

#create fq_het_id in count file
counts2$fq_het_id<-paste(counts2$fqid,counts2$position,sep="_")

#remove duplicated sequences from count file
counts2<-counts2[-which(counts2$fqid%in%dups$V1 | counts2$fqid%in%dups.33$fqid_old),]

#create fam_het_id in count file to extract all frequencies from families
counts2$fam_het_id<-paste(counts2$FID,counts2$position,sep="_")

# #check which rows are missing family info after the merge
unique(counts3$fqid[(which(is.na(counts3$FID)=="TRUE"))])

# [1] "M117-bl"            "M117-ch"            "M117C1-bl"          "M117C1-ch"          "M211-bl"            "M211-ch"           
# [7] "M211C1-bl"          "M211C1-ch"          "TR1329M_S24"        "TR23S24_S24"        "TR24S15_S15"        "TR24S24_S24"       
# [13] "TR25S12_S12"        "TR25S21_S21"        "TR26S9_S9"          "TR27S3_S3"          "TR27S6_S6"          "TR28S12_S12"       
# [19] "TR28S13_S13"        "TR28S15_S15"        "TR28S16_S16"        "TR28S17_S17"        "TR28S18_S18"        "TR28S19_S19"       
# [25] "TR28S1_S1"          "TR28S24_S24"        "TR28S3_S3"          "TR28S4_S4"          "TR28S6_S6"          "TR28S7_S7"         
# [31] "TR30S17_S17"        "TR30S1_S1"          "TR31S24_S24"        "TR33S10_S9"         "TR33S11_S10"        "TR33S12_S11"       
# [37] "TR33S13_S12"        "TR33S14_S13"        "TR33S15_S14"        "TR33S16_S15"        "TR33S17_S16"        "TR33S18_S17"       
# [43] "TR33S19_S18"        "TR33S1_S1"          "TR33S20_S19"        "TR33S21_S20"        "TR33S22_S21"        "TR33S23_S22"       
# [49] "TR33S24_S23"        "TR33S2_S2"          "TR33S4_S3"          "TR33S5_S4"          "TR33S6_S5"          "TR33S7_S6"         
# [55] "TR33S8_S7"          "TR33S9_S8"          "m205G-ch_S6"        "m287G-bl-0421_S20"  "m307-bl-0330_S6"    "m307-ch-0311_S1"   
# [61] "m307c1-bl-0330_S10" "m307c1-ch-0315_S22" "m307c2-bl-0330_S23" "m475c2_S17"         "m499-bl-0330_S19"   "m499-ch-0626_S22"  
# [67] "m499c1-bl-0330_S14" "m499c2-bl-0407_S7"  "m499c2-ch-0125_S6" 


#these are either clones or samples that were removed b/c they did not have both tissues sequenced or two children
#remove these rows

counts2<-counts2[which(is.na(counts2$FID)=="FALSE"),]

#remove families which are not present in hets3 file i.e. have no heteroplasmies after cleaning
counts3<-counts2[which(counts2$fam_het_id%in%hets5$fam_het_id),]

#add column to indicate whether heteroplasmy was discovered in that sample
counts3$heteroplasmy<-NA

counts3$heteroplasmy[which(counts3$fq_het_id%in%hets5$fq_het_id)]<-"yes"
counts3$heteroplasmy[which(is.na(counts3$heteroplasmy)==TRUE)]<-"no"
counts3$tissue_het_id<-paste(counts3$tissue_id,counts3$position,sep="_")

```

Now generate some QC statistics on the heteroplasmies. 

Calculate, for each heteroplasmy, how many times in the family it was discovered as a heteroplasmy and how many times is had MAF>0.01. There are other criteria (strand bias etc.) that are used to call a site as heteroplasmic. Sometimes sites have MAF>0.01 but they have high strandbias which might be indicative of PCR errors etc. Summarize, for each site, the difference between the number of times it was called as heteroplasmic (based on multiple stringent criteria) vs how many times it showed up with MAF>0.01. Sort sites by this difference and investigate.

```{r}

#tabulate for each heteroplasmy how many times in the family it was discovered as a heteroplasmy and how many times it had MAF>0.01
dcounts3<-counts3%>%
  group_by(position)%>%
  summarize(nmaf=length(which(maf>=0.01)),
            nhets=length(which(heteroplasmy=="yes")))

#add column listing the difference between these two numbers - this could highlight 'problematic' sites
dcounts3$diff<-with(dcounts3,nmaf-nhets)


dcounts3[which(dcounts3$diff>5),]
# position nmaf nhets diff
# 92     4233   43    12   31
# 30      567    8     1    7
# 10      185   12     6    6


```

One observation from this is that sites that show a large difference between the number of times they appear with MAF>0.01 and the number of times they are called as heteroplasmic also tend to be sites we know from prior experience are problematic (e.g. 185, 316, 204, 4233). 

Now, tabulate the number of families which are heteroplasic for the same site. Heteroplasmies are not often shared across families as low-frequency heteroplasmies don't survive for many generations. These are also often problematic sites. Identify them.

```{r}

dcounts.shared<-counts3%>%
  group_by(position)%>%
  summarize(nfams.hets=length(unique(FID[which(heteroplasmy=="yes")])),
            nfams.maf=length(unique(FID[which(maf>=0.01)])))


dcounts.shared[dcounts.shared$nfams.maf>5,]
# position nfams.maf nfams.hets
# 7        152         7          7
# 10       185         6          6
# 11       189         6          6
# 15       204         8          8
# 20       215         6          6
# 25       316         6          6
# 92      4233        16         16
# 298    16093         6          6

```

Now, identify sites with a high transversion (tv) to transition (ts) ratio. The ts/tv ratio is very high in the mtDNA. Sites with more transversion to transisitions are often sites that overlap with known problematic sites. These can be due to numt contamination or due to other sequencing artifacts.

```{r}


#identify sites with high tv/ts ration
hets5$tstv<-NA
for(i in 1:nrow(hets5)){
  if(paste(hets5$major[i],hets5$minor[i])%in%c("A G","G A","C T","T C")){
    hets5$tstv[i]<-"ts"
  }else{
    hets5$tstv[i]<-"tv"
  }
}

dhets5.tstv<-hets5%>%
  group_by(position)%>%
  summarize(nts=length(which(tstv=="ts")),
            ntv=length(which(tstv=="tv")))%>%
  arrange(-ntv)

#sites that have more transversions than transitions
dhets5.tstv[which(dhets5.tstv$ntv-dhets5.tstv$nts>5),]
# position nts ntv
# 92      4233   0  19
# 42      1291   0   8
# 25       316   0   6
# 306    16182   0   6


```

Many of the same sites fit all of these criteria. We will remove all of them from downstream analysis. 

```{r}

#remove all these sites
#make list of unique sites from all three list
x<-sort(unique(c(dcounts3$position[which(dcounts3$diff>5)],
               dcounts.shared$position[dcounts.shared$nfams.maf>5],
               dhets5.tstv$position[which(dhets5.tstv$ntv-dhets5.tstv$nts>5)])))

# [1]  4233   567   185   152   189   204   215   316 16093 16182  1291

#REMOVE THESE HETEROPLASMIES
hets6<-hets5[-which(hets5$position%in%x),]

#REMOVE THESE POSITIONS FROM COUNT FILE
counts4<-counts3[-which(counts3$position%in%x),]

#write cleaned table to file
#write cleaned table to file
fwrite(counts4,
       here("Data_files/hqcounts_cleaned_conservative_09272018.txt"),
       sep="\t",
       col.names=T,
       row.names=F,
       quote=F)

#write cleaned heteroplasmy table to file
fwrite(hets6,
       here("Data_files/Fixing_heteroplasmy_table/hq_cleaned_conservative_09272018.txt"),
       sep="\t",
       col.names=T,
       row.names=F,
       quote=F)

```

Finally, let's harmonize allele frequency across family members. The count file lists the minor allele frequency for each family member for the same site. In most cases the minor allele is the same across all members but sometimes the allele can flip, especially between mothers and children because of drift during the bottleneck. Identify such cases and list, in a separate column, the allele frequency for the same allele across all family members. This is important for downstream analyses. 

To do this, first create a table of 'reference' alleles for each site in a family based on the heteroplasmy table. This reference table will be used to adjust the allele frequency for each sample accordingly. Then write the adjusted allele frequency table to file. 

```{r}

#dhets is a table which contains the 'reference' major and minor alleles
#write function to generate major and minor alleles for each FID, position combo
f_refalleles<-function(x){
  major.list<-x$major[which(x$heteroplasmy=="yes")]
  major<-sample(major.list,1)
  minor.list<-x$minor[which(x$heteroplasmy=="yes" & x$major==major)]
  minor<-sample(minor.list,1)
  return(data.table(major=major,minor=minor))
}

dhets<-counts4%>%
  group_by(FID,position,fam_het_id)%>%
  do(f_refalleles(.))

#define function to adjust allele frequency of all individuals in a family based on this reference
adj.af<-function(x){
  fid<-as.character(x$FID[1])
  position<-x$position[1]
  major<-as.character(dhets$major[which(dhets$FID==fid &dhets$position==position)])
  minor<-as.character(dhets$minor[which(dhets$FID==fid &dhets$position==position)])
  same<-paste(major,minor,sep="")
  flip1<-paste(minor,major,sep="")
  flip2<-paste(minor,".",sep="")
  lost<-paste(major,".",sep="")
  x$adj.f<-NA
  for(i in 1:nrow(x)){
    if(paste(x$major[i],x$minor[i],sep="")%in%c(same,flip1,flip2,lost)){
      if(paste(x$major[i],x$minor[i],sep="")==same){x$adj.f[i]<-x$maf[i]}
      if(paste(x$major[i],x$minor[i],sep="")%in%c(flip1,flip2)){x$adj.f[i]<-(1-x$maf[i])}
      if(paste(x$major[i],x$minor[i],sep="")==lost){x$adj.f[i]<-0}  
    }
    if(!(paste(x$major[i],x$minor[i],sep="")%in%c(same,flip1,flip2,lost))){
      x$adj.f[i]<-0
      #if(x$major[i]==major & x$maf[i]< 0.01){x$adj.f[i]<-0}
      #else{x$adj.f[i]<-NA}
    }
  }
  return(x)}

#go through each family and 'fix' the allele frequency according to the reference alleles
hq.counts<-counts4%>%
  group_by(FID,position)%>%
  do(adj.af(.))

# hq.counts<-ddply(counts4,.(FID,position),adj.af)

#check whether this has been done correctly - the plot should look like half of an X - no off-diagonals should be observed
ggplot(hq.counts,aes(maf,adj.f))+geom_point()+theme_bw()

fwrite(hq.counts,here("Data_files/hq_counts_adj_frequency_09272018.txt")
       ,sep="\t",
       col.names=T,
       row.names=F,
       quote=F)

```


